\vspace{-1.75cm}Ph.D. Proposal

### 2022-05-17 abstract/abstract.tex
continuo abstract

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=15mm, bottom=15mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[tickmarkheight=0.7ex]{todonotes}
\newcommand{\gdv}[1]{\todo[backgroundcolor=blue]{\textbf{GDV} #1}}
\newcommand{\pb}[1]{\todo[backgroundcolor=red]{\textbf{PB} #1}}
\newcommand{\dc}[1]{\todo[[backgroundcolor=yellow]{\textbf{DC} #1}}
\usepackage[style=numeric, sorting=none, doi=false, isbn=false, url=false,
eprint=false,maxcitenames=2, mincitenames=1, maxbibnames=2, minbibnames=2,
giveninits=true]{biblatex} 
\addbibresource{abstract.bib}

@@

computational biology. This concept is based on the idea we need to move

@@

in a large collection of sequences of species. This new reference is called a

@@

field of biomedical and personalized medicine, mainly thanks to the
\textbf{genome-wide association studies (\textit{GWAS})} since GWAS studies need
to analyze a large collection of individual genomes.  Unfortunately, to complete
these types of studies on a large amount of data, such data 
must be indexed, queried and analyzed. Just to give some examples the
\textit{homo sapiens reference genome (GRCh38.p14)}, 
has a size of $\sim 3.1$\textit{Gb} and contains $\sim 59,265$ genes, as
reported by NCBI. Biological studies done with the \textit{1000 Genome Project}

@@

is necessary to consider that the objective is the sequencing of at least 100
thousand individuals, in the next few years. Therefore, it is clear that dealing
with  all these data is challenging for the current state of art algorithms and
data structures for sequence analysis. 
An example is represented by a classical research topic in sequence comparison:
\textit{pattern matching}, where the input is a long text (a

@@

occurrences of the pattern as substrings of the text.  
The problem of \textit{pattern matching} is one of the most
fundamental topics in 
the field of algorithmics and bioinformatics. We can notice that such problem
has linear solution in time, if there is proper indexing of the text. The
interest in such 

@@

patterns within the \textit{DNA}. 

@@

modeled. Among these, one of the most used is the \textbf{Burrows-Wheeler
  transform (\textit{BWT})} together with indexing via \textit{FM-index}. Thanks
to these algorithms and to the fact that it is returned an interval over the
\textit{BWT}, the problem of pattern matching, over an indexed text, has become 
linear on the length of the pattern. The
use of such 
algorithms is essential in speeding alignment algorithms such as
\textbf{BLAST}, based on the \textit{seed-and-extend paradigm}, where short
strings, the so-called \textit{seed}, are chosen to be the starting point of

@@

algorithm. The key point is the ability to compress large texts and query the
compressed texts themselves.\\  
Furthermore, in recent years, as introduced, there has been a change of interest

@@

researchers are deepening the topic of the \textbf{pangenome}, which

@@

representation of multiple genomes, encoding the variations in a multitude of
samples from the same species. In fact, the need to take into 
account the high variability in population genomes as well as the specificity of
an individual genome in a personalized approach to medicine is rapidly pushing
the abandonment of the traditional paradigm of using a single reference genome
% \cite{pangenome}
\cite{pancon}. \\

@@

of pattern matching algorithms and data structures to manage biological
sequences, with a focus on spatial 
complexity. In fact, the design of data structures to handle large
collections of data, filling the gap between the genomic data production and the
current computational state of the art, will have an impact on the ability of
molecular biologists to analyze the current amount of data, represented by
pangenome graphs. 
% \pb{the main focus of this project proposal is...}
% \dc{Corretto}
% \pb{IO menzionerei il tempo di calcolo del pattern matching del pattern ad un
% testo che è lineare nel pattern se vi è una adeguata rappresentazione o
% indicizzazione del testo}
% \dc{Aggiunto}
% \pb{il tempo che si vuole avere per ricercare tutte le occorrenze del pattern
% con la BWT è lineare nel pattern anche se il numero di occorrenze potrebbe
% essere più alto perchè si restituisce un intervallo...in altre parole tieni
% presente che dovresti dare motivazioni all'uso dell'FM-index, tenere compresso
% il testo e nel contempo avere tutte le occorrenze come intervallo di
% BWT... dire qualcosa tipo che la dimensione dei dati richieste tenere il dato
% in forma compressa e quindi poter fare le query direttamente sul dato
% compresso...puoi anche dire che lo sviluppo e il disegno di strutture succinte
% per trattare grandi collezioni di dati e la loro implementazione e
% applicazione a sequenze reali (esempio aplotipi) è uno dei temi focus del tuo
% progetto...}
% \dc{Aggiunto in modo sintetico}
% \pb{tenuto conto che il focus è filling  the current gap between the genomic
% data production and the ability of computer scintests to manage such data by
% developing new  data structure and algorithms that have to deal with these
% data. This improvement will have an impact on the ability of molecolar
% biologists to analyze the current amount of data represented by pangenome
% graphs as I will discuss in the specific goals of my project...}
% \dc{Aggiunto}
\subsubsection*{Preliminaries}
Thanks to the last developments in sequencing technologies, with \textbf{Next
  Generation Sequencing (\textit{NGS})} and \textbf{third-generation
  sequencing}, which had led both to reduce the costs of single sequencing and
to produce sequences of ever-higher quality in less and less time, the
researchers were able to theorize the \textbf{pangenome graph}. Furthermore, the
new amount of sequences has led to new algorithms for \textit{pattern matching}
for collections of sequences. In 2021, Rossi et al. proposed \textit{MONI} as a
data structure to handle a \textbf{run-length encoded version of BWT
  (\textit{RLBWT})}, with the ultimate intention of indexing and using multiple
genomes as a reference \cite{moni}. With this type of encoding, it is possible
to further improve the compression of the text, having that consecutive
identical characters, the so-called \textit{runs}, are compactly stored, making
it possible to study pangenomic sequences. According to Rossi et al.,
\textit{BWA-MEM}, one of the most used read aligners, uses between 1.1 and 3.8
times more memory than \textit{MONI}. Together with this data structure, the
authors used the concept of \textbf{matching statistics (\textit{MS})}
to efficiently compute the matches between a pattern and a text. A recent
improvement has been made through the implementation of \textit{PHONI}
\cite{phoni}, where the \textbf{longest-common-extension (\textit{LCE})
  queries}, achieving the same results as \textit{MONI} with only a single sweep
over \textit{MS} array instead of two as in \textit{MONI} (details in the next
section). 
On the other side, several computational problems arising from computational
biology have led to the development and implementation of algorithms and data
production of efficient drugs, in a context of high pharmacological resistance.
One of the most important data structures, developed to handle the
study of haplotypes sequences, is the \textbf{positional Burrows-Wheeler
  transform (\textit{PBWT})}, proposed by Durbin in 2014
\cite{pbwt_durbin}. \textit{PBWT} 
aims to represent efficiently an haplotypes panel, storing them in a queryable
compressed way (details will be described later). According to Durbin, the used
memory by \textit{PBWT} is nearly six times smaller than the raw data. These
spatial results are necessary to process real information. 
% \pb{qua dovresti essere più specifico sull'obiettivo di ridurre la memoria

@@

%\dc{I dettagli sono stati messi nello stato dell'arte}
Using this particular data structure, it is
possible to study only biallelic panel, over an alphabet $\Sigma\in\{0,1\}$, for
example, to extract common patterns in a set of haplotypes. Thanks to
the compressed representation of the 
data, the use of the \textit{PBWT} is found in many 
software for the study of haplotypes and in various genotype imputation methods
that infer unobserved genotypes in a sample of individuals \cite{impute5}.
%\pb{idem qua ...se citi anche il risparmio di memoria in sintesi può essere
%utile}
%\dc{Aggiunto il fattore di riduzione indicato da Durbin nel paper}

@@

@@

%\pb{questa parte io la imposterei con titoli diversi che evidenziano da quali}
%problemi aperti dello stato dell'arte parti per poi sviluppare gli obiettivi

@@

% Efficient and compact boolean data structures and applications...
% Inizierei dicendo che una struttura dati estremamente utile per algoritmi di
% analisi sequenze è quello dei bitvectors utilizzati tra l'altro anche per
% implementare Bloom filters...sono compatti etec... magari anche dire chi ha
% fatto SDSL in che anno ...} 
% \dc{Titoli modificati come indicato}

@@

% \dc{Non so se ho lo spazio per introdurre i bloom filters}

@@

% \pb{Run-length encoding and succinct data structures... potrebbe essere il
% titolo generale qua...tieni presente che il run lengh encoding è una vecchia
% idea usata nella compressione dei dati alternativa a LZ.. Lempel Ziv...ed è
% stata estesa alla BWT...} 

@@

type of sequences. In 2005 M\"{a}niken and Navarro defined 

@@

perspective, the algorithms have changed from being linear over the length of
the text, $n$, to be linear over the number of runs, $r$, so sub-linear over
the length of the text. The new indexing method, introduced by Gagie et al., was
called \textbf{r-index} 

@@

at the beginning and the end of every run. The algorithm for querying through
the \textit{RLBWT} takes advantage of other methods, such as the use of
\textbf{thresholds}, defined as the minimum \textit{LCP} value between two
consecutive runs of the same character, in \textit{MONI}. Instead, in
\textit{PHONI}, the authors use the \textbf{longest common extension
  (\textit{LCE}) query}, to compute the \textit{MS}. Formally, an \textit{LCE
  query}, given two positions $i$ and $j$ in a text $T$, such that $|T|=n$,
compute the length of the longest common prefix between $T[i:n-1]$ and
$T[j:n-1]$, so an \textit{LCE} is the right equal common extension 
between two positions in the text. Both solutions operate on a compressed
representation of the text, via a
\textbf{straight-line program (\textit{SLP})}. Briefly, an \textit{SLP} is based
on a \textit{grammar-compression} algorithm, by a \textit{context-free grammar},
and here it is used for \textit{random access} and \textit{LCE queries}. 
The purpose of the two projects is the computation of the \textbf{matching
  statistics (\textit{MS})}. Given a pattern $P$ and a text $T$, the \textit{MS}
of $P$ in respect to $T$ is an array $M$ of pairs position/length ($pos$/$len$),
$|M|=|P|$, such that $T[M[i].pos:M[i].pos+M[i].len-1]=P[i:i+M[i].len]-1]$ and
$P[i:i+M[i].len]$ does not occur in $T$. Given \textit{MS}, we can compute every
\textbf{Maximal Exact Match (\textit{MEM})} of a pattern in a text. Given a
text $T$ and a pattern $P$, a substring of the pattern $P[i : i+l-1]$, of length
$l$, is a \textit{MEM} of $P$ in $T$ if $P[i:i+l-1]$ is a substring of $T$ but
neither $P[i-1:i+l-1]$ nor $P[i:i+l]$ are, so if the substring cannot be
extended either to the right or to the left. Furthermore, using a particular

@@

in $SA$, the previous and next index stored in the suffix array.

@@

So, it was possible to perform pattern matching
a pangenomic context.
\paragraph*{PBWT.}
Based on the theories of the \textit{BWT}, in 2014, Durbin devised the
\textbf{positional Burrows–Wheeler transform (\textit{PBWT})}, to
solve the problem of pattern matching on panels (matrices) of haplotypes. In
detail, he analyzed a panel $X$ with $M$ haplotypes and $N$ biallelic
sites. This data structure is based on a \textit{reversed-prefix ordering} at

@@

from a new sequence $z$ to $X$ in $\mathcal{O}(M^2N)$.

@@

\textbf{IMPUTE5}, proposed by Rubinacci et al. in 2020. The main purpose of this

@@

\textit{GWAS} on tumors and other disorders. 

@@

  grow, as well as the number of sample(even if there is only a 2\% presence of
  triallelic sites, actually known in the human genome). Moreover, not only

@@

  Talking about spatial complexity, the run-length encoded version of the stored

@@

  \item \textbf{Algorithms for pangenome graphs \& other goals}. Another research
  goal is to deepen 
  the study of \textit{pangenome graphs}. In fact, it is possible to interpret

@@

  improvement would be spatial optimization. There are a lot of open problems 

@@

  \textit{BIAS} laboratory's topics of interest.

@@

@@

complete from an algorithmic point of view). Another point of interest is the

@@

@@

% \pb{nella bibliografia il lavoro MONI non è forse apparso a recomb22? non hai
% una referenza più recente?}
% \dc{non trovo la ref ufficiale a recomb21, eventualmente la posso scrivere a
% mano ma ho aggiornato con la ref a Journal of Computational Biology del 2022} 
% \pb{non metterei la referenza di Natural Computing...ma metterei quella di
% Briefings in bioinformatics sul Pangenome consortium che trovi in Natural
% computing..}
% \dc{non sono certo di aver trovato la reference indicata}
% \bibliographystyle{abbrv}
% \bibliography{abstract}

@@

% \newpage
% \noindent
% \begin{shaded}
%   \noindent
%   \textbf{Note}:
%   \begin{itemize}
%     \item il font dovrebbe essere corretto secondo le richieste. Non ho visto
%     specifiche relative ai margini del file
%     \item nell'introduzione mancano alcuni concetti relativi allo
%     studio diretto del grafo pangenomico. La professoressa Bonizzoni aveva
%     citato ad esempio le \textit{superbubbles}. Una volta chiariti gli scopri
%     precisi 
%     del progetto di ricerca dovrò sistemare (oltre che aggiungere eventuali
%     aspetti non considerati e toglierne altri)
%     \item nell'introduzione devo capire come ordinare meglio i concetti
%     \item nell'introduzione forse servono esempi di casi d'uso relativi alle
%     varie problematiche. Probabilmente è necessaria una parte più discorsiva
%     relativa anche all'importanza biologica di questo genere di studi
%     \item la conclusione dell'introduzione, dove si parla di cosa verrà
%     eventualmente approfondito durante il Ph.D., è praticamente un
%     \textit{placeholder}  
%     \item l'incipit allo stato dell'arte è sicuramente da modificare
%     \item la struttura dell'intera sezione riguardante lo stato dell'arte penso
%     debba essere rivista in base a standard per la stesura della proposal che
%     non conosco
%     \item nello stato dell'arte la parte relativa alla BWT è commentata in
%     quanto, pur avendola scritta, non la ritengo essenziale (soprattutto avendo
%     a che fare con un limite di 4 pagine)
%     \item nello stato dell'arte la sezione relativa ad alcuni esempi di studio
%     sugli aplotipi deve essere profondamente rivista in luce soprattutto dei
%     research goals. Un esempio potrebbe essere la citazione, ad esempio, di
%     HapCol, qualora i goals si spostassero un po' dalla PBWT e dai suoi usi più
%     immediati 
%     \item per ignoranza mia su come vada scritto questo tipo di documento non

>>>>

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=15mm, bottom=15mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[tickmarkheight=0.7ex]{todonotes}
\newcommand{\gdv}[1]{\todo[backgroundcolor=green]{\textbf{GDV} #1}}
\newcommand{\pb}[1]{\todo[backgroundcolor=red]{\textbf{PB} #1}}
\newcommand{\dc}[1]{\todo[backgroundcolor=yellow]{\textbf{DC} #1}}
\usepackage[style=numeric, sorting=none, doi=false, isbn=false, url=false,
eprint=false,maxcitenames=2, mincitenames=1, maxbibnames=2, minbibnames=2,
giveninits=true]{biblatex} 
\addbibresource{main.bib}

@@

computational biology based on the idea that we need to move

@@

in a large collection of individual genomes. This new reference is called a

@@

field of biology and personalized medicine, mainly thanks to 
\textbf{genome-wide association studies (\textit{GWAS})}, which need
to analyze a large collection of individual genomes.  Moreover, 
these studies also require such data  to be indexed, queried and analyzed. Just
to give some examples, from a Computer Science point of view, the 
\textit{homo sapiens reference genome (GRCh38.p14)} consists of $\sim 3.1$
characters \textit{Gb} and contains $\sim 59,265$ genes, as 
reported by NCBI. Biological studies based on the \textit{1000 Genome Project}

@@

is necessary to consider that the final goal is the sequencing of at least 100
thousand individuals, in the next few years, also thanks to latest developments
in sequencing technologies, (\textbf{Next Generation Sequencing
  (\textit{NGS})} and \textbf{third-generation sequencing}).
\dc{NGS etc citati nell'introduzione ma potrebbero essere superflui}
Therefore, it is
clear that dealing with this amount of data is challenging for the current
state-of-the-art algorithms. The problem of \textit{pattern matching} is one of
the most fundamental topics in 
the field of algorithmics and bioinformatics, where the input is a long text (a

@@

occurrences of the pattern as a substring of the text.  
We can notice that such problem
has practical and efficient linear-time solutions, if there is a proper indexing
of the text. The interest in such 

@@

patterns within a database of \textit{DNA} sequences. 

@@

designed. Among these, one of the most used is the \textbf{Burrows-Wheeler
  transform (\textit{BWT})} on which the \textit{FM-index} is based. 
%Thanks
%to these algorithms and to the fact that it is returned an interval over the
%\textit{BWT}, the problem of pattern matching, over an indexed text, has become 
%linear on the length of the pattern. 
The use of such 
algorithms is essential in speeding up alignment algorithms such as
\textbf{BLAST}, one of the most widely used aligners and implementation of one
of the most cited theoretical studies in the field of bioinformatics.
% \gdv{spiegare la rilevanza di blast} 
%\dc{BLAST ha moltissime citazioni ed è uno dei top 4 aligner}. 
This aligner based on the \textit{seed-and-extend paradigm}, where pattern
matching occurrences, the so-called \textit{seed}, are chosen to be the starting
point of 

@@

algorithm compared, for example, to those based on dynamic programming.
%\dc{confronto con DP etc}
The key point, talking about \textit{BWT}, is the ability to compress large
texts and query the compressed texts themselves, allowing alignment even 
in very large databases such as those used by \textit{BLAST}.\\
%\dc{Essenziale compressione dei db su cui lavora blast}.\\  
In recent years, there has been an important development

@@

researchers are focusing on the study of \textbf{pangenomes}, which

@@

representation of multiple genomes, encoding the variations of a multitude of
individual genomes from the same species. From a computational point of view,
\textit{pangenome} is often represented as a \textit{pangenome graph} but also
as a \textit{collection of haplotypes} \cite{pancon}, for which space complexity
can be optimised thanks to \textit{succinct data structures}. By
\textit{pangenome}, it has been possible to take into account the high
variability in population genomes, in 
addition to the specificity of single genome, allowing even more the use of
computational techniques for a \textit{personalized approach to medicine},
approach that will become increasingly important even in near future.\\ 
%In fact, the need to take into 
%account the high variability in population genomes as well as the specificity of
%an individual genome in a personalized approach to medicine is rapidly pushing
%towards leaving the traditional paradigm based on a single reference genome
%\dc{paragrafo confuso aggiungere miglioramenti tramite ds succinte e
%collegamento poi con PBWT} 

@@

of pattern matching algorithms and data structures to represent and query
biological sequences, with a focus on space 
complexity. In fact, the design of data structures and algorithms to handle
large collections of data will fill the current gap that exists between the
exponential growth of genomic data and the available state of the art algorithms
to manage  
such data. These research field will have an impact on the ability of
molecular biologists to analyze the current amount of data to extract important
genomic information from pangenome graphs. 
%\pb{the main focus of this project proposal is...}
%\pb{IO menzionerei il tempo di calcolo del pattern matching del pattern ad un
%testo che è lineare nel pattern se vi è una adeguata rappresentazione o
%indicizzazione del testo}
%\pb{il tempo che si vuole avere per ricercare tutte le occorrenze del pattern
%con la BWT è lineare nel pattern anche se il numero di occorrenze potrebbe
%essere più alto perchè si restituisce un intervallo...in altre parole tieni
%presente che dovresti dare motivazioni all'uso dell'FM-index, tenere compresso
%il testo e nel contempo avere tutte le occorrenze come intervallo di
%BWT... dire qualcosa tipo che la dimensione dei dati richieste tenere il dato
%in forma compressa e quindi poter fare le query direttamente sul dato
%compresso...puoi anche dire che lo sviluppo e il disegno di strutture succinte
%per trattare grandi collezioni di dati e la loro implementazione %applicazione
%a sequenze reali (esempio aplotipi) è uno dei temi focus del tuo progetto...} 
%\pb{tenuto conto che il focus è filling  the current gap between the genomic
%data production and the ability of computer scintests to manage such data by
%developing new  data structure and algorithms that have to deal with these
%data. This improvement will have an impact on the ability of molecolar
%biologists to analyze the current amount of data represented by pangenome
%graphs as I will discuss in the specific goals of my project...}
%\subsubsection*{Preliminaries}
%\dc{Spostare su definizioni di MS, LCE, MEM,; etc...}
%\dc{Spiegare meglio pannello di aplotipi come un pangenoma con tutti genomi di
%stessa lunghezza} 
%\dc{Per risparmiare spazio non so se l'introduzione nei preliminari della RLBWT
%  com'è ora poss andare bene} 
% The latest developments in sequencing technologies, with \textbf{Next
%   Generation Sequencing (\textit{NGS})} and \textbf{third-generation
%   sequencing}, have greatly reduced the cost of sequencing data and
% improved the quality of the data produced, the
% researchers were able to theorize the \textbf{pangenome graph}. Furthermore, the
% new amount of sequences has led to new algorithms for \textit{pattern matching}
% for collections of sequences. 
% \gdv{Questo paragrafo mi sembra una ripetizione di quanto detto nell'introduzione.
% Direi di toglierlo}
% A simple but effective strategy to compress texts is based on encoding a
% \textbf{run} of consecutive identical characters as the pair (characters, length
% of the run). For example, the string \texttt{AAAAAA} is encoded as (\texttt{A},
% 6). 
% Clearly, such an encoding achieves a better compression ratio the more
% consecutive character we can find. 
% This fact is coupled with the use of BWT, which is designed to obtain long run
% on repetitive texts, such as a pangenome, to obtain the \textbf{run-length
%   encoded BWT (\textit{RLBWT})}. 
% In fact, some recent papers --- \textit{MONI}~\cite{moni} and
% \textit{PHONI}~\cite{phoni} --- present two practical indexing tools that are
% tailored for pangenomes. 
%%In 2021, Rossi et al. proposed  as a
%data structure to handle a \textbf{run-length encoded version of BWT
%(\textit{RLBWT})},
%with the ultimate intention of indexing and using multiple
%genomes as a reference \cite{moni}, by run-length compression. 
%With this type of encoding, it is possible
%to further improve the compression of the text, having that consecutive
%identical characters, the so-called \textbf{runs}, are compactly stored, making
%it possible to study pangenomic sequences. 
%A recent
%improvement has been made through the implementation of \textit{PHONI}
%\cite{phoni}. 
%\gdv{Si riesce a dare un indizio del legame fra MEM, MS, LCE?}
%According to Rossi et al.,
%\textit{BWA-MEM}, one of the most used read aligners, uses between 1.1 and 3.8
%times more memory than \textit{MONI}. 
%Together with this data structure, the
%authors used the concept of \textbf{matching statistics (\textit{MS})}
%to efficiently compute the matches between a pattern and a text. A recent
%improvement has been made through the implementation of \textit{PHONI}
%\cite{phoni}, where the \textbf{longest-common-extension (\textit{LCE})
%  queries}, achieving the same results as \textit{MONI} with only a single sweep
%over \textit{MS} array instead of two as in \textit{MONI} (details in the next
%section). 
% On the other hand, several computational problems arising from computational
% biology have led to the development and implementation of algorithms and data
% structures for analyzing specific sequence data such as \textit{haplotypes} and
% \textit{genotypes} in the \textbf{genotyping variants problem}.
% Briefly, we
% could define \textit{haplotypes} as a combination of allelic variants, each one
% inherited from a parent. Instead, the \textit{genotype} is combined information
% of the haplotypes. For example, humans have two haplotypes, being diploid, and
% the genotype is their combined information. So, since 2005, publication of the
% \textbf{GWAS} has begun. The goal of this type of studies is to screen the
% \textit{pangenome} looking for associations between genetic variants, for
% example to study the outcomes of diseases. In this particular historical
% period, it is impossible not to mention viruses. In fact, by nature, viruses
% replicate a lot but often in a not perfect way, during infections. This produces
% many inexact clones, referred as \textbf{viral haplotypes}, which, taken
% together, form the \textbf{viral pangenome}. The identification of all these
% haplotypes is crucial both to the study of the spread of viruses and to the
% production of efficient drugs, in a context of high pharmacological resistance.
% One of the most important data structures, developed to handle the
% study of haplotypes sequences, is the \textbf{positional Burrows-Wheeler
%   transform (\textit{PBWT})}, proposed by Durbin in 2014
% \cite{pbwt_durbin}. \textit{PBWT} 
% aims to represent efficiently an haplotype panel with bi-allelic sites, storing
% them in a queryable 
% compressed way (details will be described later).
% According to Durbin, the used
% memory by \textit{PBWT} is nearly six times smaller than the raw data.
%\pb{qua dovresti essere più specifico sull'obiettivo di ridurre la memoria

@@

% From a computational point of view, he aim is to study panels built on an
% alphabet $\Sigma\in\{0,1\}$, for example, to extract common patterns in a set of
% haplotypes, computing the so-called \textit{blocks} (maximal intervals of
% columns, not extendable in any direction, such that there is a substet of
% identical rows).
%\dc{Ricordando quanto detto dalla prof Bonizzoni ho citato molto brevemente il
%concetto di blocco} 
% As introduced, haplotypes panel can be thought of as a particular case of 
% \textit{pangenome}, with fixed-length genomes, so we can understand the
% importance of such compressed and efficient data structure in computational
% pangenomics. 
%\dc{Quest'ultima parte devo sistemarla meglio}
% In addition, recent papers, such as \textit{IMPUTE5} \cite{impute5}, are based
% on the use of \textit{PBWT} for inference of unobserved genotypes in a sample of
% individuals. 
% In addition, the use of the \textit{PBWT} is found in many 
% software for the study of haplotypes and in various genotype imputation
% methods that infer unobserved genotypes in a sample of individuals
% \cite{impute5}. 
% \pb{va spegato meglio}
%\pb{idem qua ...se citi anche il risparmio di memoria in sintesi può essere utile}

@@

% \gdv{Questa sezione deve essere fusa con preliminaries, in modo da evitare
%   ripetizioni. Al momento in entrambe le sezioni si parla di RLBWT, MONI.}

@@

%\pb{questa parte io la imposterei con titoli diversi che evidenziano da quali
%%problemi aperti dello stato dell'arte parti per poi sviluppare gli obiettivi

@@

%Efficient and compact boolean data structures and applications...
%Inizierei dicendo che una struttura dati estremamente utile per algoritmi di
%analisi sequenze è quello dei bitvectors utilizzati tra l'altro anche per
%implementare Bloom filters...sono compatti etec... magari anche dire chi ha
%fatto SDSL in che anno ...} 

@@

\dc{Non so se ho lo spazio per introdurre i bloom filters}

@@

%\pb{Run-length encoding and succinct data structures... potrebbe essere il
%titolo generale qua...tieni presente che il run lengh encoding è una vecchia
%idea usata nella compressione dei dati alternativa a LZ.. Lempel Ziv...ed è
%stata estesa alla BWT...}
\dc{Non so se parlare di LZ}

@@

type of sequences and a simple but effective strategy to compress texts is based
on encoding a \textbf{run} of consecutive identical characters as the pair
(characters, length of the run). For example, the string \texttt{AAAAAA} is
encoded as (\texttt{A}, 6). 
Clearly, such an encoding achieves a better compression ratio the more
consecutive character we can find. In 2005 M\"{a}niken and Navarro defined 

@@

perspective, some recent papers --- \textit{MONI}~\cite{moni} and
\textit{PHONI}~\cite{phoni} --- present two practical indexing tools that are
tailored for pangenomes. In these works, algorithms have changed from being
linear over the length of the text, $n$, to be linear over the number of runs,
$r$, so sub-linear over the length of the text.
The new indexing method, introduced by Gagie et al., was called \textbf{r-index} 

@@

at the beginning and the end of every run. \\
To better describe how \textit{pattern matching} works in these papers, we need
to introduce some definitons. First of all, the main purpose of these algorithms
is to compute \textbf{Maximal Exact Matches (\textit{MEMs})} of a pattern in a
text. Given a text $T$ and a pattern $P$, a substring of the pattern $P[i :
i+l-1]$, of length $l$, is a \textit{MEM} of $P$ in $T$ if $P[i:i+l-1]$ is a
substring of $T$ but neither $P[i-1:i+l-1]$ nor $P[i:i+l]$ are, that is the
match cannot be extended in any direction. The authors state that computing
\textit{MEMs} is equivalent to computing \textbf{matching statistics
  (\textit{MS})}. Formally, the \textit{MS} of a pattern $P$ in respect to a
text $T$ is an array $M$ of pairs position/length ($pos$/$len$), $|M|=|P|$, such
that $T[M[i].pos:M[i].pos+M[i].len-1]=P[i:i+M[i].len]-1]$ and $P[i:i+M[i].len]$
does not occur in $T$. To compute \textit{MS}, \textit{random-access} over the
text is required  but it is also required that the text is stored in a
compressed way. For this purpose, the authors get a compressed 
representation of the text, via a \textbf{straight-line program
(\textit{SLP})}, that is based on a \textit{grammar-compression} algorithm 
(by a \textit{context-free grammar}). To compute \textit{MS}, in \textit{MONI},
the authors used the so-called \textbf{thresholds}, defined as the minimum
\textbf{Longest Common Prefix (\textit{LCP})} value between two 
consecutive runs of the same character This solution requires two sweep over the
\textit{MS array} and \textit{random access} over the \textit{SLP}. The
improvement in \textit{PHONI} is due to the fact that it was decided to use
\textbf{longest common extension (\textit{LCE}) queries}, to compute
\textit{MS}, by a single sweep. Formally, given two positions $i$ and $j$ in a
text $T$, such that 
$|T|=n$, an \textit{LCE query} compute the length of the longest common prefix
between $T[i:n-1]$ and $T[j:n-1]$. So, \textit{LCE} is the right equal common
extension between two positions in the text. Summarizing, after computing the
\textit{MS array}, as stated by Bannai et al., it is possible to obtain
\textit{MEMs}. Furthermore, using a particular

@@

in $SA$, the previous and next index stored in the suffix array. \\
Thanks to these works, it is
possible to perform \textit{pattern matching} 
a pangenomic context. In fact, for example, according to Rossi et al.,
\textit{BWA-MEM}, one of the most used \textit{read aligners}, uses between 1.1
and 3.8 times more memory than \textit{MONI}, having further improved results in
\textit{PHONI}.  
% Briefly, thanks to \textit{LCE
%   queries}, we can compute efficiently \textit{MS array}, calculating both
% position and length in a single sweep. After computing \textit{MS} we have a
% kind of representation of the \textit{MEMs} themselves.\\
%  The algorithm for querying through
% the \textit{RLBWT} takes advantage of other methods, such as the use of
% \textbf{thresholds}, defined as the minimum \textbf{Longest Common Prefix
%   (\textit{LCP})} value between two 
% consecutive runs of the same character, in \textit{MONI}, and \textit{LCE
% queries} in \textit{PHONI}, achieving the same results with only a single sweep
% over \textit{MS} array instead of two as using the \textit{thresholds}. 
%Instead, in
%\textit{PHONI}, the authors use the \textbf{longest common extension
%  (\textit{LCE}) query}, to compute the \textit{MS}. Formally, an \textit{LCE
%  query}, given two positions $i$ and $j$ in a text $T$, such that $|T|=n$,
%compute the length of the longest common prefix between $T[i:n-1]$ and
%$T[j:n-1]$, so an \textit{LCE} is the right equal common extension 
%between two positions in the text. 
%Both solutions operate on a compressed representation of the text, via a
%\textbf{straight-line program (\textit{SLP})}. Briefly, an \textit{SLP} is based
%on a \textit{grammar-compression} algorithm, by a \textit{context-free grammar},
%and here it is used for \textit{random access} and \textit{LCE queries}. 
%Given a pattern $P$ and a text $T$, the \textit{MS}
%of $P$ in respect to $T$ is an array $M$ of pairs position/length %($pos$/$len$),
%$|M|=|P|$, such that $T[M[i].pos:M[i].pos+M[i].len-1]=P[i:i+M[i].len]-1]$ and
%$P[i:i+M[i].len]$ does not occur in $T$.
%Given a
%text $T$ and a pattern $P$, a substring of the pattern $P[i : i+l-1]$, of length
%$l$, is a \textit{MEM} of $P$ in $T$ if $P[i:i+l-1]$ is a substring of $T$ but
%neither $P[i-1:i+l-1]$ nor $P[i:i+l]$ are, so if the substring cannot be
%extended either to the right or to the left.

@@

\paragraph*{Genotyping variants problems and GWAS.}
% \paragraph*{PBWT.}
Several computational problems, arising from computational
biology, have led to the development and of algorithms and data
production of efficient drugs, in a context of high pharmacological
resistance.\\
In 2014, Durbin proposed the
\textbf{positional Burrows–Wheeler transform (\textit{PBWT})} to
solve the problem of \textit{pattern matching} on panels of haplotypes,
with efficient compression. According to the author, the used
memory by \textit{PBWT} is nearly six times smaller than the raw data.
In detail, he analyzed a biallelic-sites panel $X$, built on an
alphabet $\Sigma\in\{0,1\}$, with $M$ haplotypes/individuals and $N$ sites. This
data structure is based on a \textit{reversed-prefix ordering} at 

@@

from a new sequence $z$ to $X$ in $\mathcal{O}(M^2N)$. Another interesting use
of the \textit{PBWT} is to extract common patterns in a set of
haplotypes, computing the so-called \textit{blocks} (maximal intervals of
columns, not extendable in any direction, such that there is a substet of
identical rows).

@@

\textbf{IMPUTE5} \cite{impute5}, proposed by Rubinacci et al. in 2020. The main
purpose of this 

@@

\textit{GWAS} on tumors and other disorders.\\
To concludes, as introduced,
haplotypes panel can be thought of as a particular case of  
\textit{pangenome}, with fixed-length genomes, so we can understand the
importance of such compressed and efficient data structure in
\textit{computational pangenomics}. 

@@

  grow, as well as the number of individuals (even if there is only a 2\%
  presence of triallelic sites, actually known in the human genome)
  %\pb{spiegare in sintesi cosa è il sample che viene menzionato anche sotto}.
  Moreover, not only 

@@

  Talking about space complexity, the run-length encoded version of the stored

@@

  \item \textbf{Data structures and algorithms for the representation and
    indexing  of pangenome graphs}. The above mentioned  research 
  goals are relevant in the context of a more general goal which is to deepen 
  the study of data structures for  representing the information of
  \textit{pangenome graphs}. In fact, it is possible to interpret 

@@

  improvement would be space optimization. There are a lot of open problems 

@@

  \textit{BIAS} laboratory's topics of interest. So far there has been talk of
  \textit{bitvector-based} and \textit{BWT-based indexing}. There are other data
  structures that would be interesting to study and extend, such as
  \textit{wavelet tree}, another \textit{succint data structure} used to
  compress texts. The design of new indexing structures, both for strings and
  graphs, could lead to very important results in the field of computational
  pangenomics, for example regarding the \textit{string-graph alignment problem}.
  \dc{Eventualmente penso si possa aggiungere qualcosa sull'allineamento tra
    stringhe e il grafo del pangenoma... ma non so quanto si finisca "off-topic"
    rispetto a quanto discusso nei research goals} 

@@

%\dc{Mancano lib per pangenoma (anche lato benchmark)}

@@

complete from an algorithmic point of view). Always from an experimental point
of view, there is currently a lack of libraries, in any programming language,
dedicated to the study of \textit{pangenome}, also with regard to benchmarking.
It is therefore my interest to contribute to the development of such open
source libraries. Another point of interest is the

@@

\gdv{Secondo me bisogna estendere la parte di ricerca, mettendo qualcosa di più
  a lungo termine. Ad esempio pensare a tecniche di indicizzazione diverse, non
  necessariamente basate su BWT e bitvectors} 
\dc{Allo stato attuale abbiamo una pagina di goals. Non so se possa bastare}

@@

@@
